---
title: "HW1_problem_2"
author: "Huibin Chang"
date: "2025-08-30"
output: pdf_document
---

# Problem 2: Customer churn 

```{r all, echo=TRUE, message=FALSE, warning=FALSE}
# ---- Setup ----
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(scales)
```

## 2.1 Data cleaning
I kept all the columns although the analysis in this problem is on Churn and Senior status.
For both SeniorCitizen and Churn, I created an 0-1 integer variable for ease of analysis.
```{r}
# ---- 2.1 Data preprocessing ----
telcom_churn <- read.csv("./data/Telco_Customer_Churn.csv", header=TRUE, na.strings=c("","NA"))
glimpse(telcom_churn)
#telcom_churn %>% summarize(across(everything(), ~sum(is.na(.)), .names = "na_{.col}"))

telcom_churn <- telcom_churn %>%
  mutate(
    Churn = factor(Churn, levels = c("No","Yes")),
    SeniorCitizen = if (is.numeric(SeniorCitizen)) {
      factor(SeniorCitizen, levels = c(0,1), labels = c("No","Yes"))
    } else {
      factor(SeniorCitizen, levels = c("No","Yes"))
    },
    senior_int = case_when(
      SeniorCitizen %in% c(1, "1", "Yes", "yes", TRUE) ~ 1L,
      SeniorCitizen %in% c(0, "0", "No",  "no",  FALSE) ~ 0L,
      TRUE ~ NA_integer_
    ),
    churn_int = case_when(
      Churn %in% c("Yes","yes", 1, "1", TRUE) ~ 1L,
      Churn %in% c("No", "no",  0, "0", FALSE) ~ 0L,
      TRUE ~ NA_integer_
    )
  )

telcom_churn_clean <- telcom_churn %>%
  filter(!is.na(senior_int), !is.na(churn_int))

# Sanity checks 
summary(telcom_churn_clean$Churn)
summary(telcom_churn_clean$SeniorCitizen)

telcom_churn_clean %>%
  summarise(
    n = n(),
    seniors = sum(senior_int == 1),
    non_seniors = sum(senior_int == 0),
    churn_rate_overall = mean(churn_int)
  )
```


\newpage

## 2.2 Probability of custumer churn
```{r}
# ---- 2.2 Probability of customer churn ----
n <- nrow(telcom_churn_clean)
n_churn <- sum(telcom_churn_clean$churn_int)
pi_hat <- n_churn / n
alpha <-  0.05
q <- qnorm((1 - alpha/2))
var_hat <- (pi_hat * (1 - pi_hat)) / n
se <- sqrt(var_hat)
ci_lower <- pi_hat - q * se
ci_upper <- pi_hat + q * se
# Also using other methods to verify my own calculation.
binom.test(n_churn, n)$conf.int
prop.test(n_churn, n, correct = TRUE)$conf.int

```

## 2.2 Comment
The following code chunk shows that the probability (MLE estimator) $\hat{\pi}$ is `r round(pi_hat, 3)` with a 95% CI between `r round(ci_lower, 3)` and `r round(ci_upper, 3)`. Because the CI does not cover 0, $\hat{\pi}$ is therefore **statistically** different from 0. We can of course carry out a 2-sided hypothesis testing and the conclusion is the same.



\newpage
## 2.3 Bar plots
I created 2 bar plots, one showing counts and the other percentage of churn for each group. Based on these plots, there is a visually obvious difference between the senior and non-senior groups in terms of churn.

```{r}

# ---- 2.3 Plots (two-bar summary) ----
library(ggplot2)

ggplot(telcom_churn_clean, aes(x = SeniorCitizen, fill = Churn)) +
  geom_bar() +
  labs(title = "Counts: Churn by Senior Status", x = "Senior Citizen", y = "Count")


churn_by_senior <- telcom_churn_clean %>%
  group_by(senior_int) %>%
  summarise(
    n = n(),
    churn_rate = mean(churn_int),
    .groups = "drop"
  ) %>%
  mutate(group = if_else(senior_int == 1L, "Senior", "Non-senior"))

churn_by_senior

ggplot(churn_by_senior, aes(x = group, y = churn_rate)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = percent(churn_rate, accuracy = 0.1)),
            vjust = -0.3, size = 3.8) +
  scale_y_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Churn rate by senior status", x = NULL, y = "Churn rate")


```

\newpage
## 2.4 Contingency table
```{r}
# ---- 2.4 Tables + group rates  ----
ct_table <- with(telcom_churn_clean, table(
  `Senior status` = ifelse(senior_int == 1L, "Senior", "Non-senior"),
  `Churn status`  = ifelse(churn_int  == 1L, "Yes",    "No")
))
#ct_table
#prop_by_row <- prop.table(ct_table, margin = 1)
#prop_by_row
ct_table_sum <- addmargins(ct_table)

group_rates <- telcom_churn_clean %>%
  group_by(senior_int) %>%
  summarise(
    n          = n(),
    churn_yes  = sum(churn_int == 1L),
    churn_rate = churn_yes / n,
    .groups    = "drop"
  ) %>%
  mutate(group = if_else(senior_int == 1L, "Senior", "Non-senior")) %>%
  select(group, n, churn_yes, churn_rate)
#group_rates

rate_senior <- group_rates$churn_rate[group_rates$group == "Senior"]
rate_non    <- group_rates$churn_rate[group_rates$group == "Non-senior"]
diff_rate   <- rate_senior - rate_non
sprintf("Senior churn rate: %.2f%% | Non-senior: %.2f%% | Difference: %.2f percentage points",
        100*rate_senior, 100*rate_non, 100*diff_rate)
ct_table_sum
```
## 2.4 Comment
Based on the contingency table and the probabilities, there is a pratical difference bewtween the 2 groups in churn.

\newpage
## 2.5 two confidence intervals
```{r}
# ---- 2.5 Wald & Agresti–Caffo ----
ct_table
count_senior_churn <- ct_table["Senior", "Yes"]
count_senior_nchurn <- ct_table["Senior", "No"]
count_nsenior_churn <- ct_table["Non-senior", "Yes"]
count_nsenior_nchurn <- ct_table["Non-senior", "No"]

n_seniors <- count_senior_churn + count_senior_nchurn
n_nseniors <- count_nsenior_churn + count_nsenior_nchurn
n_churns <- count_senior_churn + count_nsenior_churn
n_nchurns <- count_senior_nchurn + count_nsenior_nchurn

stopifnot(n_seniors + n_nseniors == sum(ct_table))
stopifnot(n_churns  + n_nchurns  == sum(ct_table))

pi_hat_senior <- count_senior_churn / n_seniors
pi_hat_nsenior <- count_nsenior_churn / n_nseniors
pi_hat_senior; pi_hat_nsenior

pi_diff_hat <- pi_hat_senior - pi_hat_nsenior
pi_diff_hat

pi_diff_hat_se <- sqrt((pi_hat_senior * (1 - pi_hat_senior)) / n_seniors +
                         (pi_hat_nsenior * (1 - pi_hat_nsenior)) / n_nseniors) 
pi_diff_hat_se

wald_pi_diff_ci_lower <- pi_diff_hat - q * pi_diff_hat_se
wald_pi_diff_ci_upper <- pi_diff_hat + q * pi_diff_hat_se

pi_hat_senior_ac <- (count_senior_churn + 1) / (n_seniors + 2)
pi_hat_nsenior_ac <- (count_nsenior_churn + 1) / (n_nseniors + 2)
pi_diff_hat_ac <- pi_hat_senior_ac - pi_hat_nsenior_ac
pi_diff_hat_ac

pi_diff_hat_ac_se <- sqrt((pi_hat_senior_ac * (1 - pi_hat_senior_ac)) / (n_seniors + 2) +
                         (pi_hat_nsenior_ac * (1 - pi_hat_nsenior_ac)) / (n_nseniors + 2))
ac_pi_diff_ci_lower <- pi_diff_hat_ac - q * pi_diff_hat_ac_se
ac_pi_diff_ci_upper <- pi_diff_hat_ac + q * pi_diff_hat_ac_se

ac_pi_diff_ci_upper - ac_pi_diff_ci_lower 
wald_pi_diff_ci_upper - wald_pi_diff_ci_lower 

# Also using existing method to verify my own calculation
prop.test(x = c(count_senior_churn, count_nsenior_churn),
          n = c(n_seniors, n_nseniors),
          correct = TRUE)

#tibble::tibble(
#  metric = "Diff (Senior - Non-senior)",
#  point  = pi_diff_hat,
#  Wald_L = wald_pi_diff_ci_lower, Wald_U = wald_pi_diff_ci_upper,
#  AC_L   = ac_pi_diff_ci_lower,   AC_U   = ac_pi_diff_ci_upper
#)
```
## 2.5 Comment
The MLE estimates for senior churn probability is `r round(pi_hat_senior, 3)` and `r round(pi_hat_nsenior, 3)` for non-seniors. The estimate for difference is therefore
`r round(pi_diff_hat, 5)`. 

The Wald CI ($\alpha = 0.05$) is between `r round(wald_pi_diff_ci_lower, 5)` and `r round(wald_pi_diff_ci_upper, 5)`.

For Agresti-Caffo CI, the adjusted difference (according to the textbook and the async) is `r round(pi_diff_hat_ac, 5)`, and the Agresti-Caffo CI is between `r round(ac_pi_diff_ci_lower, 5)` and `r round(ac_pi_diff_ci_upper, 5)`.

\newpage
## 2.6 Hypothesis testing
```{r}
# ---- 2.6 Hypothesis testing ----
z_value <- pi_diff_hat / pi_diff_hat_se
z_value
p_value <- 2 * (1 - pnorm(abs(z_value)))
p_value
p_value < 0.001
# So even given a significance level of (1 - 0.001 = 99.9%),
# We can reject the null that there is no difference.
``` 
## 2.6 Comment
Given the null hypothesis, the test statistic is `r round(z_value, 2)` and using the limiting distribution (Gaussian), its
corresponding (two-sided) p-value is `r round(p_value, 8)`. Because the test statistic is far beyond the critical value, we reject the null in favor of the alternative.


\newpage
## 2.7 Relative risk (rr)
```{r}
# ---- 2.7 Relative risk ----
rr_hat <- pi_hat_senior / pi_hat_nsenior 
rr_hat
var_log_rr <- (1 / count_senior_churn) - (1 / n_seniors) + (1 / count_nsenior_churn) - (1 / n_nseniors)
se_log_rr <- sqrt(var_log_rr)
log_rr_ci_lower <- log(rr_hat) - q * se_log_rr
log_rr_ci_upper <- log(rr_hat) + q * se_log_rr
rr_ci_lower <- exp(log_rr_ci_lower)
rr_ci_upper <- exp(log_rr_ci_upper)
```
## 2.7 Comment
The relative risk is `r round(rr_hat, 3)`, with a 95% CI (calculated first using log odds then exponentiate) between `r round(rr_ci_lower, 3)` and `r round(rr_ci_upper, 3)`. 

That is the senior group is 177% as likely as non-senior group to churn, or that the senior group is 77% more likely to churn compared to the non-senior group. 

This is consistent with findings from previous parts where the senior churn estimate is `r round(pi_hat_senior, 3)` and the the non-senior churn estimate is `r round(pi_hat_nsenior, 3)`

\newpage
## 2.8 Odds ratio
```{r}
# ---- 2.8 Odds ratio  ----
odds_senior <- pi_hat_senior / (1 - pi_hat_senior)
odds_nsenior <- pi_hat_nsenior / (1 - pi_hat_nsenior)
odds_ratio <- odds_senior / odds_nsenior
var_log_or <- (1 / count_senior_churn) + (1 / count_senior_nchurn) + (1 / count_nsenior_churn) + (1 / count_nsenior_nchurn)
se_log_or <- sqrt(var_log_or)
log_or_ci_upper <- log(odds_ratio) + q * se_log_or
log_or_ci_lower <- log(odds_ratio) - q * se_log_or
or_ci_upper <- exp(log_or_ci_upper)
or_ci_lower <- exp(log_or_ci_lower)
```
## 2.8 Comment
Odds (MLE) for senior to churn: `r round(odds_senior, 3)`. 

Odds (MLE) for non-senior to churn: `r round(odds_nsenior, 3)`.

The odds ratio (MLE) is  `r round(odds_ratio, 3)`.

The 95% CI for odds ratio is (calculated first using log then exponentiate) between `r round(or_ci_lower, 3)` and `r round(or_ci_upper, 3)`.